EuPPollNet: A European database of plant-pollinator networks

### 2024-08-21 Manuscript/EuPPollNet_Ms.qmd
Changes following co-authors comments

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent as defined by the European Environment Agency [@stanners1995]. As this database is the result of one of the work packages of the European project Safeguard (Safeguarding European wild pollinators; https://doi.org/10.3030/101003476), first, data was directly requested from members of the Safeguard project. Second, the request was extended to data owners outside of the project. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching for studies on Google Scholar of under-represented regions within the database. To maintain high quality standards that will support robust future ecological research, we only included studies that met the following criteria: 1) studies containing time- and geo-referenced records of plant-pollinator interactions; and 2) studies that quantify interactions by documenting them as the contact between a pollinator and the reproductive structure of a specific plant being sampled (i.e., phyto-centric networks).

@@

The database contains `r study_number` independent published and unpublished studies conducted during the time period `r min_year` - `r max_year` in `r country_number` different countries (**Figure 1a** and **Figure 1b**). The studies differ in sampling effort and methodology, and thus documenting sampling methods and sampling effort is an important feature of EuPPollNet. Most studies took place within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%), sampled a given location for an average of `r format(round(average_sampling_days, 2), nsmall=2)` days, and documented interactions mostly by using `r  most_common_method[[1]]` as the sampling method (`r format(round(most_common_method[[3]], 2), nsmall = 2)`%). All the studies documented interactions with Hymenopterans (with `r format(round(all_hymnoptera_percent, 2), nsmall = 2)`% considering all Hymenopterans, `r format(round(bee_percent, 2), nsmall = 2)`% only wild bees and `r format(round(bumblee_percent, 2), nsmall = 2)`% only bumblebees), `r format(round(Diptera_percent_metaweb, 2), nsmall = 2)`% documented interactions with Dipterans (with `r format(round(all_diptera_percent, 2), nsmall = 2)`% considering all Dipterans, `r format(round(syrphid_percent, 2), nsmall = 2)`% only syrphids and `r format(round(other_diptera_percent, 2), nsmall = 2)`% recorded syrphids plus bombylids or tachinid flies), `r format(round(Lepidoptera_percent_metaweb, 2), nsmall = 2)`% with Lepidopterans, and `r format(round(Coleoptera_percent_metaweb, 2), nsmall = 2)`% with Coleopterans. The database includes a total of `r prettyNum(interaction_number, big.mark=",")` distinct interactions. Most of the pollinator species belong to the orders Hymenoptera, Diptera, Lepidoptera and Coleoptera (`r format(round(spp_percent_main_order, 2), nsmall = 2)`%), which account for nearly all interactions in this database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). Species that belong to other orders `r format(round(100-spp_percent_main_order, 2), nsmall = 2)`% are not explored in this study as they conduct a minor fraction of the total interactions (`r format(round(100-main_orders_percent, 2), nsmall = 2)`%). Hymenoptera and Diptera contain the highest number of species comprising each approximately 1,000 species in the database. However, the majority of plant-pollinator interactions are from Hymenoptera species (`r format(round(Hymenoptera_interactions, 2), nsmall = 2)`%; **Figure 1c**). Notably, the western honey bee, _Apis mellifera_, represents `r format(round(apis_percent, 2), nsmall = 2)`% of the total interaction records from the database and an average of `r format(round(apis_average_network, 2), nsmall = 2)`% of the total interactions per network.

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only European flowering plants and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included some exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies by using their species checklists at the European level. The number of pollinator species at the European level was estimated by first estimating coverage for the three taxa that have checklists (bees, syrphids and butterflies). The mean coverage across these three groups (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`) was then used for the other taxa for which there is no checklist to extrapolate the number of pollinating species in Europe.

>>>>

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent as defined by the European Environment Agency [@stanners1995]. As this database is the result of one of the work packages of the European project Safeguard (Safeguarding European wild pollinators; https://doi.org/10.3030/101003476), first, data was directly requested from members of the Safeguard project in May 2022. Second, the request was extended to data owners outside of the project, with data collection concluding in August 2024. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching for studies on Google Scholar of under-represented regions within the database. While Google Scholar lacks reproducibility [@gusenbauer2020], it still remains the most comprehensive search engine to date [@gusenbauer2019]. This approach maximized the potential number of studies that could be incorporated in this database. The search strings used were "*plant-pollinator interactions*" and "*plant-pollinator networks*". To maintain high quality standards that will support robust future ecological research, we only included studies that met the following criteria: 1) studies containing time- and geo-referenced records of plant-pollinator interactions; and 2) studies that quantify interactions by documenting them as the contact between a pollinator and the reproductive structure of a specific plant being sampled (i.e., phyto-centric networks).

@@

The database contains `r study_number` independent published and unpublished studies conducted during the time period `r min_year` - `r max_year` in `r country_number` different countries (**Figure 1a** and **Figure 1b**). The studies differ in sampling effort and methodology, and thus documenting sampling methods and sampling effort is an important feature of EuPPollNet. Most studies took place within a single flowering season (`r format(round(one_year, 2), nsmall = 2)`%), sampled a given location for an average of `r format(round(average_sampling_days, 2), nsmall=2)` days, and documented only diurnal interactions, with `r  most_common_method[[1]]` being the most common sampling method (`r format(round(most_common_method[[3]], 2), nsmall = 2)`%). All the studies documented interactions with Hymenopterans (with `r format(round(all_hymnoptera_percent, 2), nsmall = 2)`% considering all Hymenopterans, `r format(round(bee_percent, 2), nsmall = 2)`% only wild bees and `r format(round(bumblee_percent, 2), nsmall = 2)`% only bumblebees), `r format(round(Diptera_percent_metaweb, 2), nsmall = 2)`% documented interactions with Dipterans (with `r format(round(all_diptera_percent, 2), nsmall = 2)`% considering all Dipterans, `r format(round(syrphid_percent, 2), nsmall = 2)`% only syrphids and `r format(round(other_diptera_percent, 2), nsmall = 2)`% recorded syrphids plus bombylids or tachinid flies), `r format(round(Lepidoptera_percent_metaweb, 2), nsmall = 2)`% with Lepidopterans, and `r format(round(Coleoptera_percent_metaweb, 2), nsmall = 2)`% with Coleopterans. The database includes a total of `r prettyNum(interaction_number, big.mark=",")` distinct interactions. Most of the pollinator species belong to the orders Hymenoptera, Diptera, Lepidoptera and Coleoptera (`r format(round(spp_percent_main_order, 2), nsmall = 2)`%), which account for nearly all interactions in this database (`r format(round(main_orders_percent, 2), nsmall = 2)`%). Species that belong to other orders `r format(round(100-spp_percent_main_order, 2), nsmall = 2)`% are not explored in this study as they conduct a minor fraction of the total interactions (`r format(round(100-main_orders_percent, 2), nsmall = 2)`%). Hymenoptera and Diptera contain the highest number of species comprising each approximately 1,000 species in the database. However, the majority of plant-pollinator interactions are from Hymenoptera species (`r format(round(Hymenoptera_interactions, 2), nsmall = 2)`%; **Figure 1c**). Notably, the western honey bee, _Apis mellifera_, represents `r format(round(apis_percent, 2), nsmall = 2)`% of the total interaction records from the database and an average of `r format(round(apis_average_network, 2), nsmall = 2)`% of the total interactions per network.

@@

To assess the completeness of plant and pollinator species in the EuPPollNet database at the European level, we used the aforementioned checklists for plants and pollinators. Specifically for plants, we refined the checklist to include only plants occurring in Europe and excluded taxonomic groups not associated with biotic pollination. We did this by first excluding the families considered to have exclusively a wind pollination mode [see @culley2002], and then by filtering out the genera with wind or non-biotic pollination from families that exhibit both biotic and non-biotic pollination modes. Additionally, we manually included exotic species and added unresolved species names that were not present in the accepted names of the checklist at the current version of usage. For pollinators, we compared only the taxonomic coverage of bees, syrphids and butterflies at the European level, as these are the only taxonomic groups with species checklists available in Europe. In addition, the potential number of pollinator species in Europe was estimated by combining the total number of species of bees, syrphids, butterflies and species from other taxonomic groups. The total number of species from other taxonomic groups was extrapolated by assuming that their coverage within the database is equal to the average coverage of bees, syrphids, and butterflies (mean coverage = `r round(mean_coverage, 2)`; sd = `r round(sd_coverage,2)`).
